Meta-analytic approaches and 'new' effect sizes to account for heterogeneous effects in comparative physiology

### 2021-07-05 MS/ms.Rmd
Fixing spelling

Meta-analysis is a tool that provides comparative physiologists with powerful, unbiased and quantitatively informed answers to topical research questions by synthesising effect sizes across disparate studies. Meta-analysis' core objective and strength is to distill published research results into standardised effect sizes that can be weighted (i.e., by their inverse sampling variance), and which can be compared across broad sets of research designs, study systems and species. Estimating overall effect sizes, and understanding what drives effect variability, provides opportunities to model how organisms will respond to pressing global challenges. Despite this ambition, research designs in comparative physiology can appear, at the outset, as being vastly different to each other (e.g., using different temperatures or treatment dosages). Differences in treatments across studies has led many to believe that meta-analysis is an exercise in comparing "apples with oranges". Here, we dispel this myth by showing how standardised effects sizes can be used in conjunction with powerful multi-level meta-regression models to both account for factors driving differences across studies and make them more comparable. In addition, we formalise effect size measures that provide comparative physiologists with a means to directly make effect measures comparable, removing 'nusance heterogenity' without the need to resort to more complex statistical models that may be harder to interpret. We also describe more general approaches that can be usefully applied to a variety of different contexts to derive 'new' effects sizes and sampling variances. The effect sizes and their corresponding sampling variances we describe should help physiologists deal with common treatment differences across studies.  Using effect sizes that account for components of effect heterogenity, in combination with existing meta-analytic models, will pave the way for comparative physiologists to explore exciting new questions by making results from large-scale data sets collected from the literature more accessible and widely interpretable.

@@

Meta-analysis is always concerned with effect heterogenity. In other words, the factors that drive differences in the direction and magnitude an of effect within and across studies [@Nakagawa2017; @Gurevitch2018]. The concept of heterogenity is vitally important because it tells us how general a specific conclusion is likely to be, how much of the variance we see is the result of real biology (and not just sampling variance), and provides a way to understand what the key biological factors are leading to an effect. There are a plethera of phenomena driving effect heterogenity, some of which are intersting and others that are less so. Generally, contributors to effect size heterogenity fall within a number of broad categories. First, sampling variance contributes to heterogenity; the variability in an effect because of the inherent variability in sampling a given population. We cannot sample the entire population and nor do we necessarily want to. As a consequence, we need to rely on a sub-sample to estimate a population parameter (e.g., correlation between two traits). How many independent samples we collect impacts our precision in estimating this 'true' effect. In meta-analysis we formally include sampling variance to remove this sourec of heterogenity directly - hence why having sampling variances for an effect is so important. Second, there are socialogical factors impacting heterogenity. These include attitudes towards publishing non-significant results. Low powered studies that do not find an 'effect' or that find an effect in the opposite direction often lead to what is called publication bias. Third, there is real biology driving differences in effects. This is particularly true in comparative physiology where different species with varying habitats, life-histories, mating systems and reproductive modes are synthesised. As comparative physiologists, this is the stuff that gets us up every morning! Much variation could result from shared evolutionary history (i.e., phylogeny) or because of the different thermal environments species (or populations) occupy. We can model these biological factors to test predcitions on what we hypothsise are key players impacting an effect improving our understanding of the biologcial world around us. Finally, methodological factors are also big contributers to effect variability. In some cases, these methodological factors are of direct interest. For example, the methods for measuring an outcome variable can result in different effects, and it's improtant we know and understand this as it can determine how expetriments are designed in the future. In other cases, methodological factors might be considered 'nusance heterogenity' - factors we know vary, but are rather unsurprising to us. These can include differences in tempertature, pH, dosages, water potential etc. that are obviously going to vary and have an impact on the magnitude of an effect. In some cases, these factors might actually not be 'nusance' variables but rather the main interest. However, in many cases the can result in a more complicated interpretation of an effect measure then what is desired.

All the various sources of heterogenity (except sampling variance) all contribute (to varying extents) to the 'apples and oranges' problem often touted as being a problem in meta-analyses (or quantitaive syntheses more generally). Combinding standardised effects, while ignoring these sources of heterogenity can result in the meaning of a meta-analysis not being particularly useful to many. Many others, however, argue that this is just what we are interested in synthesizing. In the words of Gene Glass, the father of meta-analysis: "Of course it mixes apples and oranges; in the study of fruit nothing else is sensible; comparing apples and oranges is the only endeavor worthy of true scientists; comparing apples to apples is trivial" [p.g., 224, @Glass2015]. @Glass2015's intention was to simply highlight that whether comparing apples and oranges is a good idea simply boils down to the specific question of interest and the population one wishes to make conclusions about. If we are interested in gaining insight into 'fruit' then synthesising apples and oranges is the only thing that makes sense. <!---  as said, comparative analysis by definition, compares applies and oranges - = different species - I guess you are mainly talking about mixing different methods ones - this is where I said earlier - focusing on explaining hetero comes in. Also, mixing different quality of studies are often referred to as "Garbage in and garbage out" ---> Instead, if we are interested in generalising only to apples then mixing fruit will not be ideal. We tend to agree with @Glass2015, and his sentiment is of particular relevance to comparative physiology. For example, if we want to understand the impact of a pollutant on reproduction in aquatic organisms then including aquatic mammals and fish would provide answers to this general question, even if their reproductive biology and physiology are known to differ. By synthesizing diverse study outcomes, it allows us to not just report on overall effects (if the question is relevant to ask), but understand critical features of the literature, study systems (e.g., different effects of pollutants on fish and mammal reproduction) and approaches that can explain the diversity of effect outcomes observed. In doing so, it provides a richer set of conclusions that can draw attention to important sources of variability that might have otherwise been overlooked. Having said that, it is still very important to understand the limitations and diversity of studies included within a meta-analysis - an important reason why transparency and reproducibility are such a prominent feature of meta-analysis (or ideally should be) [@ODea2021]. The validity of general conclusions to certain questions (e.g., what is the overall effect of a pollutant on reproduction?) will inevitably depend on the homogeneity of effects being synthesized. <!--- this is know as putting one number on so many different kinds of studies - but of course, this may not be that interesting apart from a starting point for subsequent meta-regression analyses ---> For example, given the very different physiology of aquatic mammals and fish, understanding an overall combined effect on reproduction may not be particularly relevant, and indeed may be misleading. It is important to consider these questions when conducting meta-analyses.

Of course, there are certain types of heterogeneity that transcend different questions. In other words, if we compare different types of 'apples' is it still fair to synthesize these studies to understand overall effects? This situation might be analogous to a meta-analysis in comparative physiology that attempts to understand the effects of temperature on fish growth across different populations of a widely distributed (cosmopoliton) species. Studies might rear populations of fish under different temperatures, but in the end, they are still all the same fish species, governed by the same thermodynamic effects on growth. In this situation, we might still be interested in understanding overall effects, but we may question whether such a conclusion is sensible given the diversity of temperatures applied. It is not always clear how we should tackle these sources of heterogenity. If temperature differences are of direct interest, then we should definitely model them. If, however, it is less interesting, than say the habitat occupied then it may introduce an unnecessary complication to the interpretation of the effect(s) of interest. Regardless, a diversity of factors *a priori* that are hypothesised or known to impact heterogenity does not preclude us from also exploring all these important moderators or predictors of variability in effects across studies; however, depending on the effect size used it may just complicate their interpretation or make modelling of all these effects challenging.

@@

As discussed above, it is clear that different types of questions and experimental approaches will mix 'apples and oranges' to varying extents, and indeed, this heterogenity (or basket of fruit) is often the main interest to comparative biologists. To better understand the different types of effect sizes being used, thier susceptability to known sources of methodological / environmental variablity across studies, and how these sources are controlled we conducted a literature survey. To accomplish this goal, we searched for meta-analyses and comparative analyses that tackled questions using physiological data compiled from animal studies across multiple species. We were specifically interested in asking: 1) what type of effect size was used within the quantitative synthesis? 2) based on the methods provided, was the effect size likely susceptible to continuous variation in how treatments were applied (e.g., for contarst-based effects) or the conditions under which traits were measured (i.e., for correlatioons) across studies? 3) If so, was the variation resulting from treatment differences accounted for in the effect size or during analysis?, and finally, 4) how were effect sizes analysed in the quantitative synthesis, were these weighted analyses, and what program was used for analysis? We also recorded whether publication bias and sensitivity analyses were conducted, if data were publicly available, where it was stored, and if so, if it was usable (i.e., in a machine readable format).

@@

## **Conclusions and Future Directions**
Comparative physiologists are intersted in meta-analysing effects across a diversity of species, experimental approaches / designs and environments. While an overall meta-analytic mean may be of interest, more often than not, understanding the drivers of effect size heterogenity will be the primary interest. However, obvious drivers of effect size heterogenity, such as temperature and dosage differences across studies, may not really be of interest. Indeed, comparative physiolgists are likely to want an effect size or analytical approach that allows them to remove these 'nusances'. Here, we provide comparative physiologists with a set of tools to deal with 'nusance heterogenity' by creatively 'reinventing' standard effect sizes or using meta-regerssion models. Of course, nusance heterogenity might in of itself be the main interest, in which case standard meta-regression approaches will work sufficiently. However, in many cases, comparative physiolgists might simply want to focus on bigger, more interesting biological moderators of effect size direction and magnitude. However, estimating complex meta-regression models with a multitude of fixed effects might result in more challenging model interpretation. As such, a combined approach, where some of the known heterogenity is dealt with by deriving a variant effect size and associated sampling variance, and other biologically interesting moderators are fit in a meta-regerssin model, might be the desired. Indeed, this may even be a necessity to simplify the model.

>>>>

Meta-analysis is a tool that provides comparative physiologists with powerful, unbiased and quantitatively informed answers to topical research questions by synthesising effect sizes across disparate studies. Meta-analysis' core objective and strength is to distill published research results into standardised effect sizes that can be weighted (i.e., by their inverse sampling variance), and which can be compared across broad sets of research designs, study systems and species. Estimating overall effect sizes, and understanding what drives effect variability, provides opportunities to model how organisms will respond to pressing global challenges. Despite this ambition, research designs in comparative physiology can appear, at the outset, as being vastly different to each other (e.g., using different temperatures or treatment dosages). Differences in treatments across studies has led many to believe that meta-analysis is an exercise in comparing "apples with oranges". Here, we dispel this myth by showing how standardised effects sizes can be used in conjunction with powerful multi-level meta-regression models to both account for factors driving differences across studies and make them more comparable. In addition, we formalise effect size measures that provide comparative physiologists with a means to directly make effect measures comparable, removing 'nuisance heterogeneity' without the need to resort to more complex statistical models that may be harder to interpret. We also describe more general approaches that can be usefully applied to a variety of different contexts to derive 'new' effects sizes and sampling variances. The effect sizes and their corresponding sampling variances we describe should help physiologists deal with common treatment differences across studies.  Using effect sizes that account for components of effect heterogeneity, in combination with existing meta-analytic models, will pave the way for comparative physiologists to explore exciting new questions by making results from large-scale data sets collected from the literature more accessible and widely interpretable.

@@

Meta-analysis is always concerned with effect heterogeneity. In other words, the factors that drive differences in the direction and magnitude an of effect within and across studies [@Nakagawa2017; @Gurevitch2018]. The concept of heterogeneity is vitally important because it tells us how general a specific conclusion is likely to be, how much of the variance we see is the result of real biology (and not just sampling variance), and provides a way to understand what the key biological factors are leading to an effect. There are a plethora of phenomena driving effect heterogeneity, some of which are interesting and others that are less so. Generally, contributors to effect size heterogeneity fall within a number of broad categories. First, sampling variance contributes to heterogeneity; the variability in an effect because of the inherent variability in sampling a given population. We cannot sample the entire population and nor do we necessarily want to. As a consequence, we need to rely on a sub-sample to estimate a population parameter (e.g., correlation between two traits). How many independent samples we collect impacts our precision in estimating this 'true' effect. In meta-analysis we formally include sampling variance to remove this source of heterogeneity directly - hence why having sampling variances for an effect is so important. Second, there are sociological factors impacting heterogeneity. These include attitudes towards publishing non-significant results. Low powered studies that do not find an 'effect' or that find an effect in the opposite direction often lead to what is called publication bias. Third, there is real biology driving differences in effects. This is particularly true in comparative physiology where different species with varying habitats, life-histories, mating systems and reproductive modes are synthesised. As comparative physiologists, this is the stuff that gets us up every morning! Much variation could result from shared evolutionary history (i.e., phylogeny) or because of the different thermal environments species (or populations) occupy. We can model these biological factors to test predictions on what we hypothesize are key players impacting an effect improving our understanding of the biological world around us. Finally, methodological factors are also big contributors to effect variability. In some cases, these methodological factors are of direct interest. For example, the methods for measuring an outcome variable can result in different effects, and it's important we know and understand this as it can determine how experiments are designed in the future. In other cases, methodological factors might be considered 'nuisance heterogeneity' - factors we know vary, but are rather unsurprising to us. These can include differences in temperature, pH, dosages, water potential etc. that are obviously going to vary and have an impact on the magnitude of an effect. In some cases, these factors might actually not be 'nuisance' variables but rather the main interest. However, in many cases the can result in a more complicated interpretation of an effect measure then what is desired.

All the various sources of heterogeneity (except sampling variance) all contribute (to varying extents) to the 'apples and oranges' problem often touted as being a problem in meta-analyses (or quantitative syntheses more generally). Combining standardised effects, while ignoring these sources of heterogeneity can result in the meaning of a meta-analysis not being particularly useful to many. Many others, however, argue that this is just what we are interested in synthesizing. In the words of Gene Glass, the father of meta-analysis: "Of course it mixes apples and oranges; in the study of fruit nothing else is sensible; comparing apples and oranges is the only endeavor worthy of true scientists; comparing apples to apples is trivial" [p.g., 224, @Glass2015]. @Glass2015's intention was to simply highlight that whether comparing apples and oranges is a good idea simply boils down to the specific question of interest and the population one wishes to make conclusions about. If we are interested in gaining insight into 'fruit' then synthesising apples and oranges is the only thing that makes sense. <!---  as said, comparative analysis by definition, compares applies and oranges - = different species - I guess you are mainly talking about mixing different methods ones - this is where I said earlier - focusing on explaining hetero comes in. Also, mixing different quality of studies are often referred to as "Garbage in and garbage out" ---> Instead, if we are interested in generalising only to apples then mixing fruit will not be ideal. We tend to agree with @Glass2015, and his sentiment is of particular relevance to comparative physiology. For example, if we want to understand the impact of a pollutant on reproduction in aquatic organisms then including aquatic mammals and fish would provide answers to this general question, even if their reproductive biology and physiology are known to differ. By synthesizing diverse study outcomes, it allows us to not just report on overall effects (if the question is relevant to ask), but understand critical features of the literature, study systems (e.g., different effects of pollutants on fish and mammal reproduction) and approaches that can explain the diversity of effect outcomes observed. In doing so, it provides a richer set of conclusions that can draw attention to important sources of variability that might have otherwise been overlooked. Having said that, it is still very important to understand the limitations and diversity of studies included within a meta-analysis - an important reason why transparency and reproducibility are such a prominent feature of meta-analysis (or ideally should be) [@ODea2021]. The validity of general conclusions to certain questions (e.g., what is the overall effect of a pollutant on reproduction?) will inevitably depend on the homogeneity of effects being synthesized. <!--- this is know as putting one number on so many different kinds of studies - but of course, this may not be that interesting apart from a starting point for subsequent meta-regression analyses ---> For example, given the very different physiology of aquatic mammals and fish, understanding an overall combined effect on reproduction may not be particularly relevant, and indeed may be misleading. It is important to consider these questions when conducting meta-analyses.

Of course, there are certain types of heterogeneity that transcend different questions. In other words, if we compare different types of 'apples' is it still fair to synthesize these studies to understand overall effects? This situation might be analogous to a meta-analysis in comparative physiology that attempts to understand the effects of temperature on fish growth across different populations of a widely distributed (cosmopolitan) species. Studies might rear populations of fish under different temperatures, but in the end, they are still all the same fish species, governed by the same thermodynamic effects on growth. In this situation, we might still be interested in understanding overall effects, but we may question whether such a conclusion is sensible given the diversity of temperatures applied. It is not always clear how we should tackle these sources of heterogeneity. If temperature differences are of direct interest, then we should definitely model them. If, however, it is less interesting, than say the habitat occupied then it may introduce an unnecessary complication to the interpretation of the effect(s) of interest. Regardless, a diversity of factors *a priori* that are hypothesised or known to impact heterogeneity does not preclude us from also exploring all these important moderators or predictors of variability in effects across studies; however, depending on the effect size used it may just complicate their interpretation or make modelling of all these effects challenging.

@@

As discussed above, it is clear that different types of questions and experimental approaches will mix 'apples and oranges' to varying extents, and indeed, this heterogeneity (or basket of fruit) is often the main interest to comparative biologists. To better understand the different types of effect sizes being used, their susceptibility to known sources of methodological / environmental variability across studies, and how these sources are controlled we conducted a literature survey. To accomplish this goal, we searched for meta-analyses and comparative analyses that tackled questions using physiological data compiled from animal studies across multiple species. We were specifically interested in asking: 1) what type of effect size was used within the quantitative synthesis? 2) based on the methods provided, was the effect size likely susceptible to continuous variation in how treatments were applied (e.g., for contrast-based effects) or the conditions under which traits were measured (i.e., for correlations) across studies? 3) If so, was the variation resulting from treatment differences accounted for in the effect size or during analysis?, and finally, 4) how were effect sizes analysed in the quantitative synthesis, were these weighted analyses, and what program was used for analysis? We also recorded whether publication bias and sensitivity analyses were conducted, if data were publicly available, where it was stored, and if so, if it was usable (i.e., in a machine readable format).

@@

## **Conclusions and Future Directions**
Comparative physiologists are interested in meta-analysing effects across a diversity of species, experimental approaches / designs and environments. While an overall meta-analytic mean may be of interest, more often than not, understanding the drivers of effect size heterogeneity will be the primary interest. However, obvious drivers of effect size heterogeneity, such as temperature and dosage differences across studies, may not really be of interest. Indeed, comparative physiologists are likely to want an effect size or analytical approach that allows them to remove these 'nuisances'. Here, we provide comparative physiologists with a set of tools to deal with 'nuisance heterogeneity' by creatively 'reinventing' standard effect sizes or using meta-regression models. Of course, nuisance heterogeneity might in of itself be the main interest, in which case standard meta-regression approaches will work sufficiently. However, in many cases, comparative physiologists might simply want to focus on bigger, more interesting biological moderators of effect size direction and magnitude. However, estimating complex meta-regression models with a multitude of fixed effects might result in more challenging model interpretation. As such, a combined approach, where some of the known heterogeneity is dealt with by deriving a variant effect size and associated sampling variance, and other biologically interesting moderators are fit in a meta-regression model, might be the desired. Indeed, this may even be a necessity to simplify the model.
